---
title: "AARUG Workshop Takehome"
subtitle: "Old Faithful Dataset"
author: "your name here"
date: "May 21, 2016"
output: html_document
---

# Instructions

This R Markdown document is the *optional* takehome portion of the Ann Arbor R User Group workshop.

The takehome is organized as a series of tasks or questions followed by a R Markdown code chunk. Write code into the code chunk that will complete the task or answer the question. It is good practice to comment your code (using `#`). Also feel free to write additional documentation about your code or answer.

#### Example: Use R as a calculator to add 3 and 7.

```{r example}
3 + 7  # add 3 and 7
```

The answer should be 10.

```{r load packages, results='hide'}
# if you are using any packages it is good practice to load them at the beginning
```

# Problems

## Import the dataset

Download the Old Faithful dataset from the [workshop page](http://annarborrusergroup.github.io/AARUG-R-workshop/).

The `Old-Faithful.csv` dataset contains waiting times between eruptions and the duration of the eruption for the Old Faithful geyser in Yellowstone National Park, Wyoming, USA.

#### Use the read.csv function to read in the dataset and assign it to a variable named `oldfaithful`.
*Hint:* be sure the file is available from your current working directory.

```{r import data}

```

#### Show the structure of the data using the `str()` function.

```{r data structure}

```

#### Show how many rows and columns are in the dataset.

```{r dimensions}

```

## Indexing and data manipulation

#### Show rows 12-20 of the dataset using bracket indexing `[ ]`.

```{r bracket indexing}

```

#### Get the `interval` column of the dataset using bracket indexing `[ ]`.

```{r bracket indexing with characters}

```

#### Get the `dur.cat` variable out of the data using the `$` notation.

```{r `$` indexing}

```

#### How many eruptions had a short duration?

```{r number of short eruptions}

```

#### Create a new dataframe that contains only the eruptions with a short duration.

```{r subset}

```

#### What is the average (mean) duration of the eruptions in your new dataframe?

```{r average eruption duration}

```

#### What is the standard deviation of the eruptions in your new dataframe?
*Hint:* Use the `sd()` function

```{r eruption duration standard deviation}

```

## New functions

This section introduces some new functions that we didn't discuss during the workshop. Remember that you can get documentation for a function by using `?` and the function name. For example `?summary` will give you the documentation for the `summary()` function.

#### The function `table()` creates a summary table with the counts for each factor. Use `table()` to recalculate the number of short and long duration eruptions in the full dataset

```{r table}

```

#### The function `unique()` returns the unique values of an object. Use `unique()` to get unique values for the `day` column.

```{r unique}

```

#### The function `quantile()` calculates the quantiles for a numeric variable. Use `quantile()` to get the quantiles of eruption duration.

```{r quantile}

```

***
