---
title: "Manipulating Data with Tidyverse"
author: "Jyotishka Datta"
date: "5/18/2019"
output: 
  html_document:
      toc: true
      number_sections: false
fontsize: 12pt
urlcolor: blue
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

Manipulating here means moving the same data into different arrangements without changing the actual values.

In this tutorial, you'll learn how to do these things in a fast way that's easy to repeat using a set of software tools known as the "Tidyverse".

The Tidyverse, a collection of R packages built around the foundational idea that data in a table should have one variable per column, one observation per row and only one value per cell.

One of the packages in the Tidyverse is dplyr, which is a package for getting at parts of your data in a fast, easy-to-understand way that's easy to replicate. In this tutorial, we're going to learn about and practice using the six core "verbs" of data manipulation in the Tidyverse.

> ## Learning Objectives
>

> By time you finish this tutorial, you will be able to take a data frame and:

> * Use the pipe (%>%) to create a seamless workflow
> * Use select() to select one or more columns
> * Use filter() to select one or more rows
> * Use mutate() to add new variables
> * Use arrange() to change the order of rows
> * Use summarize() to convert a variable to a single value
> * Use group_by() to group sets of observations

## Getting Ready 

This will be pretty quick: we just need to read in the library we're going to use (tidyverse), then read in our data and put it in a data frame. We're going to read our data in using the read_csv() function rather than read.csv(), which will put it in a tibble.

Tibble: A "modern take on data frames", it's bascially a data frame that is designed to be more user-friendly and avoid some of the defaults that lead to frequent bugs.

For this tutorial, we're going to use an agricultural survey dataset that has information on more than 9500 families. It's not a huge dataset in terms of what you might encounter in industry, but it is large enough that interacting with it in a spreadsheet program would be a pain.

```{r}
# library we're going to use
library(tidyverse)

# read in that data
farmData <- read_csv("data/herding-cats-small.csv")
```

Now that we've got our data frame read into R, let's take a peek at the first few rows.

```{r}
# print the first ten rows of our dataframe
head(farmData, 10)
```

```{r}
# get the number of rows and columns in our dataset
dim(farmData)
```


